---
title: "Individual Project Preliminary Analyses"
author: "Hannah Rodgers"
date: "4/27/2022"
output: pdf_document
bibliography: preliminary_bibliography.bib
---

##This R script is used to clean up and transform soil data to get ready for regression analysis.

# Overview of Goals

  My research focuses on how soil health and soil microbiology influence long-term sustainability in semiarid wheat systems. Wheat is one of the most important crops worldwide, yet the semiarid landscapes where it is primarily grown are particularly vulnerable to climate change and land degradation [@asseng2015rising; @assessment2018fourth]. In particular, the traditional wheat-fallow system practiced throughout the High Plains ecoregion inefficiently stores water, depletes soil fertility, and has a high potential for erosion, pushing farmers and researchers to search for economical ways to build long-term soil health [@norton2012loss; @kaur2015effects]. Sustaining agriculture in the High Plains will require alternative farming systems that can rebuild soil health while remaining profitable [@hansen2012].

   Soil health is critical for sustainable and resilient food production, yet evaluating soil health in semiarid climates remains challenging. Agriculture impacts soil health through practices such as crop rotation, fallowing, tillage, and fertilization, though these effects vary depending on environmental factors such as climate and soil type. Soil microorganisms are sensitive indicators of changing soil health, and can help evaluate the long-term sustainability of land management practices [@rodgers2021effects]. However, there are gaps in our understanding of how soil microbial properties should be interpreted, especially in semiarid systems. 
   
   This project uses soil data from an experiment on the long-term impacts of compost application on soil health and soil microbiology in Wyoming High Plains organic wheat systems. Specifially, four rates of compost was applied 
to small plots in a randomized complete block design in either 2016 or 2021. In contrast to fresh manure or chemical fertilizers, composted manure is made of concentrated, stabilized nutrients and carbon that release slowly over time and improve soil physical properties [@larney2006]. This includes data on soil physical properties (such as aggregate stability and bulk density), soil chemical properties (such as total organic matter, labile carbon and nitrogen pools, and pH) and soil microbiology (enzyme activity, microbial biomass, and microbial community composition). I aim to use regression analysis to evaluate the relationships between compost rate and soil properties. However, many of the variables do not meet the regression assumptions. Therefore, this project focuses on cleaning and tidying my dataset, testing the regression assumptions on each variable, transforming the data to meet those assumptions, and finally visualizing the data. Ultimately, quantifying soil health benefits over many years could help growers work towards building the long-term health of their soil, and could guide efforts to protect soils and create an agricultural system resistant in the face of climate change and land degradation.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load Packages
```{r 1}
library(MASS)
library(readxl)
library(writexl)
library(patchwork)
library(ggpubr)
library(ggbiplot)
library(tidyverse)
```

#Import Data
```{r 2}
setwd("C:/Users/hanna/Desktop/GRAD PROJECTS/OREI")

#read  files
OREI_soil <- read_excel("OREI_05.2021.xlsx", sheet = "Soil_Data")
OREI_treatments <- read_excel("OREI_05.2021.xlsx", sheet = "Treatment_Data_2021")
OREI_enzymes <- read_excel("OREI_05.2021.xlsx", sheet = "Enzymes")
OREI_PLFA <- read_excel("OREI_05.2021.xlsx", sheet = "PLFAs")

#units 'cheat sheet'
units <- t(read_excel("OREI_05.2021.xlsx", sheet = "Units"))
```

#Clean Up Data
```{r 3}
#merge the sheets
OREI_all <- OREI_treatments %>% 
  left_join(OREI_soil, by = 'Sample_ID') %>% 
  left_join(OREI_enzymes, by = 'Sample_ID') %>% 
  left_join(OREI_PLFA, by = 'Sample_ID')

OREI_all <- OREI_all %>% 
#remove fallow phase and inorganic fertilized plots
  filter(Rotation == "wheat", Treatment != "fertilizer") %>% 
#remove unwanted variables
 dplyr::select(-PER1, -ID, -InorganicC, -Treatment, -Compost_Rate, 
         -Crop, -Rotation, -H2O, -BulkDensity) %>% 
#save compost year as a factor
  mutate(Compost_Year = as.factor(Compost_Year))

#separate into the two groups
OREI_2016 <- subset(OREI_all, Compost_Year != "2020")
OREI_2020 <- subset(OREI_all, Compost_Year != "2016")
```

# Remove Outliers

Outliers can skew a regression, so first I'll remove any outliers from each variable.
```{r 4}
#check each variable for outliers. Run on each variable in both OREI_2016 and OREI_2020
source("http://goo.gl/UUyEzD")
outlierKD(OREI_2020, total_bacteria)

#OREI_2016, removed outliers from: yield, NO3, PMN, DON, MBC, CBH, PHOS, NAG, actino,  gram_pos, AMF, sapro_fungi, total_MB, total_fungi

#OREI_2020, removed outliers: NO3, protein, MBC, MBN, SOC, N, NAG, BX, AG, SUL, 

#save no_outlier file to disk, then read it back in for next session
#write_xlsx(OREI_2020, "OREI_2020_no_outliers.xlsx")
#write_xlsx(OREI_2016, "OREI_2016_no_outliers.xlsx")

OREI_2020 <- as.data.frame(read_excel("OREI_2020_no_outliers.xlsx"))
OREI_2016 <- as.data.frame(read_excel("OREI_2016_no_outliers.xlsx"))
```

## CHECK REGRESSION ASSUMPTIONS #
1. Observations are independent.
2.Normality. The residuals are normally distributed.
3. Linear relationship. The relationship between X and Y is linear.
4. Homoscedasticity. The residuals have constant variance for all values of X.


#2. Check for normality of residuals. Transform any non-normal variables.
```{r}
#create an empty list
p.vals <- list()

#this loop runs a linear model on compost ~ each variable, evaluates normality with the Shapiro test, and saves the p.value
for (i in names(OREI_2016[,7:38])) {
  mod <- lm(get(i) ~ compost, data = OREI_2016)
  p.vals[[i]] <- (shapiro.test(mod$residuals))$p.value }

#any variable with p>0.05 has non-normal residuals

#OREI_2020 with non-normal: DOC, DON, PHOS
#OREI_2016: only PHOS

#this function performs the optimal box_cox transformation
box_cox_transform <- function(v) {
  
#calculates the boxcox plot and pulls out lambda
  bc <- boxcox(v ~ compost, data = OREI_2016)
  lambda <- bc$x[which.max(bc$y)]
  
#transforms the data using lambda and saves it
  v <- (v^lambda-1)/lambda
  return(v) }

#run box_cox_transform on all non-normal variables
OREI_2016$PHOS <- box_cox_transform(OREI_2016$PHOS)
OREI_2020$PHOS <- box_cox_transform(OREI_2020$PHOS)
OREI_2020$DOC <- box_cox_transform(OREI_2020$DOC)
OREI_2020$DON <- box_cox_transform(OREI_2020$DON)

#test normality of new data using same loop from above
p.vals.trans <- list()

for (i in names(OREI_2016[,7:38])) {
  mod <- lm(get(i) ~ compost, data = OREI_2016)
  p.vals.trans[[i]] <- (shapiro.test(mod$residuals))$p.value }

p.vals.trans

#everything is normal now! Save again:
#write_xlsx(OREI_2020, "OREI_2020_normal.xlsx")
#write_xlsx(OREI_2016, "OREI_2016_normal.xlsx")

OREI_2020 <- as.data.frame(read_excel("OREI_2020_normal.xlsx"))
OREI_2016 <- as.data.frame(read_excel("OREI_2016_normal.xlsx"))
```


#3. Check linearity using F-test for lack of fit.
```{r}
#this function runs a linear and quadratic model for each variable, and tests whether they differ significantly
F_test <- function(x) {
  mod <- lm(x ~ compost, data = OREI_2016)
  reduced<-lm(x ~ compost, data = OREI_2016)
  full<-lm(x ~ poly(compost,2), data = OREI_2016)
  return(anova(reduced, full)$"Pr(>F)")
}

#use sapply to run this function on each

p.vals.F.test <- list()

for (i in colnames(OREI_2016[,7:38])) {
  p.vals.F.test[[i]] <- F_test(OREI_2016[[i]])
}

#differs: 2020 DOC, protein, porosity, WFPS
#2016: all good!

#####4. check constancy of residuals (levene test? brown forsythe?) look at Liana's code ####
library(car)
leveneTest(mod, compost)
```



#### VISUALIZE DATA ####
```{r}
p1 <- ggplot(data= OREI_normal, aes(x = compost, y = DOC))+
  geom_point() +
  geom_smooth(method = 'lm') +
  stat_regline_equation() +
  labs (title = 'One Year After Compost Application', x = "Compost Rate (Mg/ha)", y = "DOC (mg/kg)")

p2 <- ggplot(data= OREI_normal, aes(compost, DOC)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_regline_equation() +
  labs (title = 'Five Years After Compost Application', x = "Compost Rate (Mg/ha)", y = "DOC (mg/kg)")

p1 + p2 #& scale_y_continuous(limits = c(0, 375)) 
```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

# References
